---
title: "Intro to making maps with R (ggplot2)"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    chakra: libs/remark.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5, warning = FALSE, message = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
duo_accent(
  primary_color = "ivory",
  secondary_color = "#310A31",
  header_font_google = google_font("Roboto", "400"),
  text_font_google   = google_font("Lato", "300"),
  code_font_family = "Fira Code",
  code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css",
  header_color = "#f54278",
  title_slide_text_color = "#354a66"
)
```


## Background: R Spatial

#### General Comments:

  * R has an extensive collection of packages and functionality for doing geospatial data analysis and geocomputation
  
  * many packages have overlapping functionality 
  
  * newer packages have built on older packages 
  
  * May be intimidating to navigate the landscape of geospatial functionality available
  
  



---
## Resources

For more extensive information about doing geospatial analysis in R, I've used:

--

  * Angela Li's R spatial workshop notes
  
    * [https://spatialanalysis.github.io/workshop-notes/](https://spatialanalysis.github.io/workshop-notes/)
    
--

  * Geocomputation with R by Robin Lovelace, Jakub Nowosad, Jannes Muenchow
  
    * [https://geocompr.robinlovelace.net/](https://geocompr.robinlovelace.net/)
    
--

Both available in bookdown versions as of 2020-03-27



---
## What we are not doing today


--

  * anything too advanced
  
--

  * talking too much about projections
  
--

  * working with raster data
  
  

---
## What we are doing today

--

  * a brief review of <tt>ggplot2</tt>
  
--

  * working with vector formats
    * shapefiles, geoJSON
    
--
  
  * using the <tt>sf</tt> package to read and store geospatial data
  

--

  * Using <tt>ggplot2</tt> to make a chloropleth map using a new geometry, <tt>geom_sf</tt>
  
  
--

  * at some point, we will use some join functionality from the <tt>dplyr</tt> package to merge geospatial data with a table of aggregate statistics
  

--


  * possibly make a second chloropleth map, if we have the time



---
## The data we will be using

#### Definitely:

--

  * Data from the Chicago Data Portal
    * Shapefile/geoJSON representations of communities in Chicago
    * Community-level public health/socioeconomic data
    * Restaurant inspection data for February 2020 
    
--

#### Maybe:

--

  * Data from Open Data Toronto
    * Shapefile of electoral districts 
    * Voter statistics from the 2018 Toronto City Council Election
    
--

#### Unlikely (but not against the laws of Physics)
  
--
  
  * Data from NYC Open Data
    * Shapefiles/geoJSON of zip codes in New York City
    * 311 logs for Jan 1 2020 to Jan 14 2020
    


---
### Geospatial data formats

  * Shapefiles
    * widely used, developed by ESRI commonly used in Arc-branded software
    * information is stored as a collection of points, lines, and polygons (and combinations of points, lines and polygons)
    * non-geospatial info can be stored as well
    * take the form of a folder/directory of associated files

--

  * geoJSON: 
    * open format for storing geospatial information (points, lines, polygons)
    * can also store non-geospatial information
    * based on JavaScript Object Notation (JSON)
  
