---
title: "Installing R and RStudio"
author: "Daryn Ramsden"
format: html
editor: visual
---



### 1. Install R and RStudio



Download R @ The R Project’s Home Page
  * Windows:
        [https://cran.r-project.org/bin/windows/base/R-4.2.1-win.exe]
  * Mac (Intel):
        https://cran.r-project.org/bin/macosx/base/R-4.1.2.pkg
  * Mac (ARM: new mac models manufactured since November 2020)
        https://cran.r-project.org/bin/macosx/big-sur-arm64/base/R-4.1.2-arm64.pkg
  * Linux https://cloud.r-project.org/bin/linux/

Download RStudio Desktop from RStudio
  * Windows https://download1.rstudio.org/desktop/windows/RStudio-2021.09.1-372.exe
  * Mac https://download1.rstudio.org/desktop/macos/RStudio-2021.09.1-372.dmg
  * Linux https://www.rstudio.com/products/rstudio/download/#download


**Run the installers for R and RStudio, in that order.**

### 2. Open RStudio

It should look like this when you do:


### 3. Install additional packages and data

  * copy-paste the text below into the command window
  * hit enter
  * accept any defaults if you get prompted
  * this may take a little while. But nothing too exorbitant.

This picture should show where you paste the command before hitting enter

```{r eval = FALSE}
install.packages(c("tidyverse", "nycflights13", "devtools"))
```




### Test it out

Copy-paste the following text into the command window and hit enter.

If the graphs show up then you have installed all the necessary packages. Don’t worry if you don’t understand the commands.


```{r}
library(tidyverse)
library(nycflights13)
flights <- flights |> 
  mutate(cancelled = is.na(arr_delay) & is.na(dep_delay),
         Day = ISOdate(year, month, day))

flight_report <- flights |>
  group_by(Day, origin) |> 
  summarise(Cancelled = mean(cancelled)*100)

ggplot(flight_report, aes(x = Day, y = Cancelled)) +
  geom_line() +
  facet_wrap(~origin, ncol = 1) +
  labs(title = "Flight cancellations in NYC 2013", 
       y = "% of Flights Cancelled")
```





You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
