---
title: "dplyr 1: key verbs"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    chakra: libs/remark.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5, warning = FALSE, message = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
duo_accent(
  primary_color = "ivory",
  secondary_color = "#310A31",
  header_font_google = google_font("Roboto", "400"),
  text_font_google   = google_font("Lato", "300"),
  code_font_family = "Fira Code",
  code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css",
  header_color = "#f54278",
  title_slide_text_color = "#354a66",
  padding = "15px 15px 25px 15px"
)
```


## dplyr: a package for data manipulation

The data you get is almost in the form you want

--

<tt>dplyr</tt> is an R package that encapsulates many common data manipulation tasks

--

Sometimes you want to: 
--

  * keep only some of the rows

--

  * keep only some of the columns

--

  * adds new columns

--

  * sort data
  
--

  * provide summary statistics

--

<tt>dplyr</tt> has a function for each of these

  
---
## 5 Key tasks/verbs/functions

  * <tt>filter</tt>: keep only some of the rows 


  * <tt>select</tt>: keep only some of the columns 

  * <tt>mutate</tt> adds new columns



  * <tt>arrange</tt>: sort data
  


  * <tt>summarise</tt> (and <tt>group_by</tt>): provide summary statistics 
  
---
## First, let's get some data

```{r message = FALSE, warning = FALSE, eval = FALSE}
library(readr)

chi_emps <- read_csv("data/ChicagoEmployees.csv")
```

---
## First, let's get some data

```{r message = FALSE, warning = FALSE, eval = FALSE}
library(readr)

chi_emps <- read_csv("data/ChicagoEmployees.csv")
chi_emps
```


---
## First, let's get some data

```{r message = FALSE, warning = FALSE}
library(readr)

chi_emps <- read_csv("data/ChicagoEmployees.csv")
chi_emps
```



---
### What is this data?


--
A list of all employees of the City of Chicago (Jan 2020)
--

#### What do the fields represent?

  * Name: employee's last name and first initial

--

  * Title: job title

--

  * Dept: Department of city government

--

  * SalHr: Whether the employee is salaried (<i>S</i>) or hourly-paid (<i>H</i>)
  
--
  
  * Hrs: Typical hours per week for an hourly-paid employee
  
--

  * Salary: Annual Salary for salaried employee

--

  * Rate: Hourly rate for hourly-paid employee
  


---
## How do we use dplyr?

--

```{r eval = TRUE}
library(dplyr)
```
--


### Or ...

--
you can load the entire <tt>tidyverse</tt> 

```{r eval = FALSE}
library(tidyverse)
```



---
### If you've never installed it before 

--

You will need to install the package first 

```{r eval = FALSE}
install.packages("dplyr")
```

--
or 

```{r eval = FALSE}
install.packages("tidyverse")
```




---
### Key thing to remember

--

All of these functions have these in common:

--

  * the first input is a data frame 
  
--

  * there are a variable number of following inputs specifying the manipulation you want to do


---
## filter

#### a function for specifying which rows to keep


Example 1: How do we get all employees that are salaried?
--
```{r message = FALSE, eval = FALSE}
salaried <- filter(chi_emps, SalHr == "S")
```

---
## filter

#### a function for specifying which rows to keep


Example 1: How do we get all employees that are salaried?

```{r message = FALSE, eval = FALSE}
salaried <- filter(chi_emps, SalHr == "S")
salaried
```

  
---
## filter

#### a function for specifying which rows to keep


Example 1: How do we get all employees that are salaried?

```{r message = FALSE}
salaried <- filter(chi_emps, SalHr == "S")
salaried
```


---
## filter

#### 


Example 2: How do we get all employees of the Police Department?
--
```{r}
police <- filter(chi_emps, Dept == "POLICE")
head(police)
```


---
## filter

#### 

Example 3: How do we get all salaried employees of the Police Department?
--
```{r}
sal_police <- filter(chi_emps, SalHr == "S", 
                     Dept == "POLICE")

sal_police
```

---
### Assessment

How many hourly-paid employees are there in the Aviation department (<i>AVIATION</i>)?


--

```{r}
aviation_hr <- filter(chi_emps, Dept == "AVIATION",
                      SalHr == "H")
dim(aviation_hr)
```





---
## select


#### a function for specifying which columns to keep

Example: How do we keep only the names, titles, departments and salaries of the salaried employees?  

```{r eval = FALSE}
salaried2 <- select(salaried, Name, Title, Dept, Salary)
```

---
## select


#### a function for specifying which columns to keep

Example: How do we keep only the names, titles, departments and salaries of the salaried employees?  

```{r eval = FALSE}
salaried2 <- select(salaried, Name, Title, Dept, Salary)
salaried2
```


---
## select


#### a function for specifying which columns to keep

Example: How do we keep only the names, titles, departments and salaries of the salaried employees?  

```{r}
salaried2 <- select(salaried, Name, Title, Dept, Salary)
salaried2
```


---
## select

#### Alternative: we can use the <tt>-</tt> to specify columns we don't want to keep:

--

```{r eval = FALSE}
salaried3 <- select(salaried, -SalHr, -Hrs, -Rate)
```


---
## select

#### Alternative: we can use the <tt>-</tt> to specify columns we don't want to keep:


```{r eval = FALSE}
salaried3 <- select(salaried, -SalHr, -Hrs, -Rate)
salaried3
```
---
## select

#### Alternative: we can use the <tt>-</tt> to specify columns we don't want to keep:


```{r}
salaried3 <- select(salaried, -SalHr, -Hrs, -Rate)
salaried3
```



---
## arrange 

### a function to sort your data

Example: How do we sort the data so that employees are ordered by salary?

```{r}
salary_ord <- arrange(salaried, Salary)
```



---
## arrange

Example: How do we sort the data so that employees are ordered by salary in descending order?

```{r}
salary_desc <- arrange(salaried, desc(Salary))
```


---
## arrange

You can also sort by more than one variable

Example: How do we sort the salaried employees by Department (alphabetically) with each department sorted by salary (descending)?


```{r}
dept_sal <- arrange(salaried, Dept, desc(Salary))
```



--

Question: How can you check that this works?

--

  * One way: Try the <tt>View</tt> function with the data frame you created as input.



---
### Assessment

Fill in the code below to do the following tasks:

  * start with the original data set (all employees)
  * restrict to hourly-paid employees
  * remove the <i>SalHr</i>, and <i>Salary</i> columns
  * order the data in descending order of hourly rate
    * if employees earn the same hourly rate order them by typical hours (descending) 
  
  
```{r eval = FALSE}
____ <- filter(____, ____ == ____)
____ <- select(____, -____, -_____)
____ <- arrange(____, ____, ____)
```




---
### Assessment solution





```{r}
df <- filter(chi_emps, SalHr == "H")
df2 <- select(df, -SalHr, -Salary)
df3 <- arrange(df2, desc(Rate), desc(Hrs))

df3
```

(You don't have to use the same data frame names.)


---
### Assessment solution: alternative


  * You could also have overwritten the data frame at each step.
    * your environment is less cluttered
    * but seeing the intermediate data may help with troubleshooting/debugging your code

```{r}
df <- filter(chi_emps, SalHr == "H")
df <- select(df, -SalHr, -Salary)
df <- arrange(df, desc(Rate), desc(Hrs))

df
```



---
## mutate

#### a function to add new columns


Example: Adding a column that indicates whether an individual has an annual salary greater than 
$100,000 to the salaried data set

```{r eval = FALSE}
sal100 <- mutate(salaried, above100 = if_else(Salary > 100000, TRUE, FALSE))
```

---
## mutate

#### a function to add new columns


Example: Adding a column that indicates whether an individual has an annual salary greater than 
$100,000 to the salaried data set

```{r eval = FALSE}
sal100 <- mutate(salaried, above100 = if_else(Salary > 100000, TRUE, FALSE))
sal100
```


---
## mutate

#### a function to add new columns


Example: Adding a column that indicates whether an individual has an annual salary greater than 
$100,000 to the salaried data set

```{r}
sal100 <- mutate(salaried, above100 = if_else(Salary > 100000, TRUE, FALSE))
sal100
```




---
## summarise

#### a function for summary statistics


Example: Provide the following summary statistics for the City of Chicago:

  * the total number of employees
  * the total salaries paid to salaried employees
  * the mean rate paid to hourly-paid employees

```{r eval = FALSE}
report <- summarise(chi_emps, NumEmps = n(),
                    TotalSalaries = sum(Salary, na.rm = TRUE),
                    MeanRate = mean(Rate, na.rm = TRUE))
```

---
## summarise

#### a function for summary statistics


Example: Provide the following summary statistics for the City of Chicago:

  * the total number of employees
  * the total salaries paid to salaried employees
  * the mean rate paid to hourly-paid employees

```{r eval = FALSE} 
report <- summarise(chi_emps, NumEmps = n(),
                    TotalSalaries = sum(Salary, na.rm = TRUE),
                    MeanRate = mean(Rate, na.rm = TRUE))
report
```

---
## summarise

#### a function for summary statistics


Example: Provide the following summary statistics for the City of Chicago:

  * the total number of employees
  * the total salaries paid to salaried employees
  * the mean rate paid to hourly-paid employees

```{r} 
report <- summarise(chi_emps, NumEmps = n(),
                    TotalSalaries = sum(Salary, na.rm = TRUE),
                    MeanRate = mean(Rate, na.rm = TRUE))
report
```



---

## summarise

#### How can we find the following for the Police Department?

  * the total number of employees
  * the median salary

--

```{r}
police_report <- summarise(police, NumEmps = n(),
                           MedSal = median(Salary, na.rm = TRUE))
```

---

## summarise

#### How can we find the following for the Police Department?

  * the total number of employees
  * the median salary



```{r eval = FALSE}
police_report <- summarise(police, NumEmps = n(),
                           MedSal = median(Salary, na.rm = TRUE))
police_report
```


---

## summarise

#### How can we find the following for the Police Department?

  * the total number of employees
  * the median salary



```{r}
police_report <- summarise(police, NumEmps = n(),
                           MedSal = median(Salary, na.rm = TRUE))
police_report
```



---


## group_by

#### A function to put the data into groups (before you summarise)


```{r}
grouped_chi <- group_by(chi_emps, Dept)
grouped_chi
```


---

#### How is the grouped data frame different from the original?
--

```{r}
grouped_chi
```


  * Notice the information about groups above the data

  * Rows have been put into 36 different groups based on their department
  


---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each department, what are:
  * the number of employees
  * the median salary of salaried employees?
  

---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each department, what are:
  * the number of employees
  * the median salary of salaried employees?
  
```{r}
grouped_chi <- group_by(chi_emps, Dept)
```
 
 
---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each department, what are:
  * the number of employees
  * the median salary of salaried employees?
  
```{r eval = FALSE}
grouped_chi <- group_by(chi_emps, Dept)
report <- summarise(grouped_chi, NumEmps = n(), 
                    MedSal = median(Salary, na.rm = TRUE))
``` 



---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each department, what are:
  * the number of employees
  * the median salary of salaried employees?
  
```{r eval = FALSE}
grouped_chi <- group_by(chi_emps, Dept)
report <- summarise(grouped_chi, NumEmps = n(), 
                    MedSal = median(Salary, na.rm = TRUE))

head(report)
```


---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each department, what are:
  * the number of employees
  * the median salary of salaried employees?
  
```{r}
grouped_chi <- group_by(chi_emps, Dept)
report <- summarise(grouped_chi, NumEmps = n(), 
                    MedSal = median(Salary, na.rm = TRUE))

head(report)
```

---
### Assessment

#### Create a data frame that:
  * contains the number of employees in each department that earn more than $120,000/yr
  * is ordered by the number of employees meeting the criterion (descending)

Reorder the following lines of code and fill in the blanks where necessary


```{r eval = FALSE}
____ <- arrange(____, desc(____))
____ <- summarise(____, ____ = n())
____ <- filter(____, ____ > ____)
____ <- group_by(____, ____)
```



---
### Assessment 

#### How would you change your previous answer to show the departments in alphabetical order?

