---
title: "An overview of dplyr"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: libs/switch-themer.css
    chakra: libs/remark.js
    nature:
      highlightLines: true
      countIncrementalSlides: false
    includes:
      after_body: libs/toggle.html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      comment = "")
xaringanExtra::use_tile_view()
#xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
#xaringanExtra::use_editable()
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE  
)

library(palmerpenguins)
library(dplyr)
library(kableExtra)
```



### The data we will be using 


```{r}
#install.packages("palmerpenguins")
library(palmerpenguins)
```

```{r echo = FALSE, eval = FALSE}
DT::datatable(penguins, 
         extensions = c('FixedColumns',"FixedHeader"),
          options = list(scrollX = TRUE, 
                         paging=TRUE,
                         fixedHeader=TRUE,
                         pageLength = 15))
```

```{r echo = FALSE}
library(rmarkdown)
paged_table(penguins)
```




---
### What do these variables represent?

Data were collected and made available by Dr. Kristen Gorman and the Palmer Station, Antarctica LTER, a member of the Long Term Ecological Research Network.

  * *species*: *Adelie*, *Chinstrap* or *Gentoo*
  
  * *island*: *Biscoe*, *Dream* or *Torgersen* (factor)
  
  * *bill_length_mm*: bill length mm (numeric) 
  
  * *bill_depth_mm*: bill depth in mm (numeric)      
  
  * *flipper_length_mm*: flipper length in mm (numeric)
  
  * *body_mass_g*: body mass in grams (numeric)
  
  * *sex*: *male* or *female* (factor)        
  
  * *year*: 2007, 2008 or 2009
  
  
---
## dplyr: a package for data manipulation

The data you get is almost in the form you want

--

<tt>dplyr</tt> is an R package that encapsulates many common data manipulation tasks

--

Sometimes you want to: 
--

  * keep only some of the rows

--

  * keep only some of the columns

--

  * adds new columns

--

  * sort data
  
--

  * provide summary statistics

--

<tt>dplyr</tt> has a function for each of these



---
## 5 Key tasks/verbs/functions

  * <tt>filter</tt>: keep only some of the rows 


  * <tt>select</tt>: keep only some of the columns 

  * <tt>mutate</tt> adds new columns



  * <tt>arrange</tt>: sort data
  


  * <tt>summarise</tt> (and <tt>group_by</tt>): provide summary statistics 

  
  
---
## `filter`

#### a function for specifying which rows to keep


Example 1: How do we get all penguins of the Chinstrap species?



---
## `filter`

#### a function for specifying which rows to keep


Example 1: How do we get all penguins of the Chinstrap species?

```{r eval = FALSE, message = FALSE}
chinstrap <- filter(penguins, species == "Chinstrap")
```


---
## `filter`

#### a function for specifying which rows to keep


Example 1: How do we get all penguins of the Chinstrap species?

```{r message = FALSE}
chinstrap <- filter(penguins, species == "Chinstrap")
chinstrap
```


---
## `filter`

#### a function for specifying which rows to keep


Example 2: How do we get penguins that are 4 kg or greater?

  
---
## `filter`

#### a function for specifying which rows to keep


Example 2: How do we get penguins that are 4 kg or greater?


```{r eval = FALSE}
penguins_4k <- filter(penguins, body_mass_g >= 4000)
```
---
## `filter`

#### a function for specifying which rows to keep


Example 2: How do we get penguins that are 4 kg or greater?


```{r}
penguins_4k <- filter(penguins, body_mass_g >= 4000)
penguins_4k
```


---
### Assessment

How many penguins were found on Torgersen island (<i>Torgersen</i>)?


  
---
### Assessment

How many penguins were found on Torgersen island (<i>Torgersen</i>)?



```{r}
torgersen<- filter(penguins, island == "Torgersen")
dim(torgersen)
```

--

Also could have used:


```{r}
torgersen<- penguins %>% filter(island == "Torgersen")
dim(torgersen)
```




---
## `select`

#### A function/verb for specifying which columns to keep 

As of dplyr 1.0 there are 5 ways to use select 

  1. By **position** 
  
  2. By **name**
  
  3. by **function of name** 
  
  4. by **type** 
  
  5. by combination of the above using logical operators (`|`, `&`, `!`)



---
### `select` by position

Example: select columns 1, 3 and 5 from `penguins`

--
```{r eval = FALSE}
penguins %>% select(1, 3, 5)
```


---
### `select` by position

Example: select columns 1, 3 and 5 from `penguins`

```{r}
penguins %>% select(1, 3, 5)
```


---
### `select` by name

Example: select *species*, *island* and *body_mass_g*

--
```{r eval = FALSE}
penguins %>% select(species, island, body_mass_g)
```

---
### `select` by name


Example: select *species*, *island* and *body_mass_g*

```{r}
penguins %>% select(species, island, body_mass_g)
```


---
### `select` by a function of column names


Example: Choose all columns that contain "mm":

```{r}
penguins_mm <- penguins %>% select(contains("mm"))
```


---
### `select` by a function of column names


Example: Choose all columns that contain "mm":

```{r}
penguins_mm <- penguins %>% select(contains("mm"))

penguins_mm
```


---
### `select` by a function of column names


Example: How to choose all columns starting with "bill":

```{r}
bills_df <- penguins %>% select(starts_with("bill"))
```

---
### `select` by a function of column names


Example: How to choose all columns starting with "bill":

```{r}
bills_df <- penguins %>% select(starts_with("bill"))

bills_df
```


---
### `select` by type 

Example: choose all numeric columns:

```{r eval = FALSE}
penguins %>% select(where(is.numeric))
```

---
### `select` by type 

Example: choose all numeric columns:

```{r}
penguins %>% select(where(is.numeric))
```


---
### `select` by logical combination

Example: choose all factor variables or variables containing the word "bill"

```{r eval = FALSE}
penguins %>% select(where(is.factor) | contains("bill"))
```



---
### `select` by logical combination

Example: choose all factor variables or variables containing the word "bill"

```{r}
penguins %>% select(where(is.factor) | contains("bill"))
```


---
### `select` by type 

```{r}
penguins %>% select(where(is.numeric))
```


---
### `select` by logical combination


```{r}
penguins %>% select(where(is.factor) | contains("bill"))

```



---
## `mutate` 

#### a function to add new columns


Example: Adding a column that indicates whether a penguin has a mass greater than 4 kg

```{r eval = FALSE}
penguin_extra <- penguins %>% 
  mutate(above_4kg= if_else(body_mass_g > 4000, TRUE, FALSE))
```


---
## `mutate` 

#### a function to add new columns


Example: Adding a column that indicates whether a penguin has a mass greater than 4 kg

```{r}
penguin_extra <- penguins %>% 
  mutate(above_4kg= if_else(body_mass_g > 4000, TRUE, FALSE))

head(penguin_extra)
```


---
## `arrange` 

#### A function for sorting data

Example: Sort all penguins by body mass:

--

```{r}
penguins_sorted <- penguins %>% arrange(body_mass_g)
```

---
## `arrange` 

#### A function for sorting data

Example: Sort all penguins by body mass:



```{r eval = FALSE}
penguins_sorted <- penguins %>% 
  arrange(body_mass_g)
penguins_sorted
```
```{r echo = FALSE}
paged_table(penguins_sorted)
```



---
### sorting with multiple columns using `arrange`

Example sorting by species, then by descending order of mass:

```{r eval = FALSE}
penguins_sorted2 <- penguins %>% 
  arrange(species, desc(body_mass_g))
penguins_sorted2
```
```{r echo = FALSE}
penguins_sorted2 <- penguins %>% 
  arrange(species, desc(body_mass_g))
paged_table(penguins_sorted2)
```





---
## `summarise`/`summarize`


#### A verb/function to get summary statistics.


Question: what's the mean flipper length and body mass among the Palmer penguins?


```{r}
penguins %>% 
  summarise(num_penguins = n(),
            avg_mass = mean(body_mass_g, na.rm = TRUE),
            avg_fl_length = mean(flipper_length_mm, na.rm = TRUE))
```


---
## `group_by` 

#### A function that makes `summarise` really powerful 

`group_by` creates a grouped data frame based on columns you specify

--

For example, grouping the penguins by island and species:

--

```{r}
gr_penguins <- penguins %>% group_by(island, species)
```


---
## `group_by` 

#### A function that makes `summarise` really powerful 

`group_by` creates a grouped data frame based on columns you specify



For example, grouping the penguins by island and species:

```{r}
gr_penguins <- penguins %>% group_by(island, species)
head(gr_penguins)
```





---
## How is the grouped data frame different?

--

  * Extra information is added to the data frame 
  
--


  * rows that match on all the grouping variables will be in the same group

--
  
  * rows that don't match on all the grouping variables will be in different groups
  


---
## `group_by` and `summarise` together

Now let's do the same summary as before with the grouped data:

--

```{r}
gr_penguins %>% summarise(num_penguins = n(),
                       avg_mass = mean(body_mass_g, na.rm = TRUE),
                       avg_fl_length = mean(flipper_length_mm, 
                                            na.rm = TRUE))
```

  
---
### So ... a couple other things about groups

  * default behavior is to remove the last level of grouping after a call to `summarise` 
  
  * grouped data can be used with other `dplyr` verbs e.g. `mutate`
  
  * you can ungroup data using `ungroup`
  
---
### Example using `group_by` with `mutate`

What if we wanted to give each penguin a number within its species? 

```{r}
numbered_penguins <- penguins %>% 
  group_by(species) %>%
  mutate(penguin_num = 1:n())
```

---
### Example using `group_by` with `mutate`

What if we wanted to give each penguin a number within its species? 

```{r}
numbered_penguins <- penguins %>% 
  group_by(species) %>%
  mutate(penguin_num = 1:n())

numbered_penguins
```



---
## `rename`

#### A function/verb to rename columns

Works like `select`


Example: renaming by position

```{r}
penguins_different <- penguins %>% rename(bill_length = 3, 
                                          bill_depth = 4)

penguins_different
```



---
### `rename_with`

`rename_with` can be used with a specified transformation (and optionally with a column selection). 


Example: rename all columns to be uppercase

```{r}
penguins %>% rename_with(toupper)
```



---
## `rename_with`
```{r}
penguins %>% rename_with(toupper, where(is.numeric))
```



---
## `relocate`

### A function 


  * (**default**) move selected variables to the front
  
  * move selected columns before a specified location
  
  * move selected columns after a specified location



---
## `relocate` examples

Example: bring all the factor variables to the front 

```{r eval = FALSE}
penguins %>% relocate(where(is.factor))
```


---
## `relocate` examples

Example: bring all the factor variables to the front 

```{r}
penguins %>% relocate(where(is.factor))
```



---
## `relocate` examples

Example: relocate all factor variables after *body_mass_g*  


```{r}
penguins %>% relocate(contains("bill"), .after = body_mass_g)
```



---
### `across`: a really useful new function

What if you wanted the average value - per group - of each numeric column? 

--

Annoying way: 

```{r}
penguins %>% group_by(species) %>% 
  summarise(avg_bill_length = mean(bill_length_mm, na.rm = TRUE),
            avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE), 
            avg_fl_length_mm = mean(flipper_length_mm, na.rm = TRUE),
            avg_body_mass_g = mean(body_mass_g, na.rm = TRUE))
```


---
### `across`: a really useful new function

What if you wanted the average value - per group - of each numeric column? 


Neater/better way: 

```{r}
penguins %>% group_by(species) %>%
  summarise(across(where(is.numeric) & !contains("year"),
                   mean, na.rm = TRUE))
```

---
### multiple summaries with `across`

Example: For each island, what is the average of all numeric variables and the count of all factor variables?

```{r}
penguins %>%
  group_by(island) %>% 
  summarise(
    across(where(is.numeric), mean, na.rm = TRUE), 
    across(where(is.factor), n_distinct),
    n = n(), 
  )
```


---
### `across` example with `filter`

Example: get all rows without missing values:


```{r}
penguins_complete <- penguins %>% 
  filter(across(everything(), ~ !is.na(.x)))
```

--

Is that any different to?

```{r eval = FALSE}
penguins_complete2 <- penguins %>% 
  filter(across(everything(), complete.cases))
```


---
### `across` example with `distinct`

All combinations of variables meeting specified criteria using `distinct`

```{r}
penguins %>% distinct(across(is.factor, sort = TRUE))
```

---
### `across` example with `count`

Counts of all combinations of variables meeting specified criteria using `count`

```{r}
penguins %>% count(across(is.factor, sort = TRUE))
```

---
### `across` example with `mutate`

Using `across` with `mutate` to rescale all numeric variables between 0 and 1


```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

penguins_rescaled <- penguins %>% 
  mutate(across(where(is.numeric), rescale01))

penguins_rescaled
```
