---
title: "Money and Payments in Trinidad and Tobago"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: libs/switch-themer.css
    chakra: libs/remark.js
    nature:
      highlightLines: true
      countIncrementalSlides: false
    includes:
      after_body: libs/toggle.html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      comment = "")
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
xaringanExtra::use_editable()

library(tidyverse)
library(readxl)
library(hrbrthemes)
library(lubridate)
library(scales)
library(gt)

thousand <- 1000
hundred <- 100
million <- 1000000
```


---
title: "Payment Systems and Money in T&T"
author: "Daryn Ramsden"
date: "last updated: `r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    highlight: haddock
    theme: readable
    css: dr-themer.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      comment = "")

library(tidyverse)
library(readxl)
library(hrbrthemes)
library(lubridate)
library(scales)
library(gt)

thousand <- 1000
hundred <- 100
million <- 1000000
```

## How can you make payments in Trinidad and Tobago

There are several payment media:

--
  
  1. Cash
  
--
  
  2. Cheques
  
--
  
  3. Electronic Payments, including:
  
--
    1. Debit Cards
    
--
    2. Credit Cards
    
--
    3. ACH: Automated Clearing House
    
--
    4. RTGS: Real Time Gross-Settlement System

---
# Currency in Active Circulation


```{r}
a <- 100
```


---
# Debit Cards

## Daily usage of debit and credit cards

```{r echo = FALSE}
debit_cards <- read_csv("data/DebitCardTransactions.csv") %>% 
  mutate(DATE = my(DATE)) %>% 
  rename(Year = YEAR)
credit_cards <- read_csv("data/CreditCardTransactions.csv") %>% 
  mutate(DATE = my(PERIOD),
         Year = year(DATE)) 
ydl <- read_csv("data/year_days_lookup.csv")

thousand <- 1000
cc_year <- credit_cards %>% 
  group_by(Year) %>% 
  summarise(`POS transactions` = sum(POSVOLUME),
            `POS Value` = sum(POSVALUE_TTM),
            `ATM transactions` = sum(ATM_VOLUME),
            `ATM value` = sum(ATMVALUE_TTM)) %>% 
  left_join(ydl) %>% 
  mutate(`CC Point of Sale transactions per day` = paste0(round(`POS transactions`/(thousand*Days)),"k"),
         `CC ATM withdrawals per day` = paste0(round(`ATM transactions`/(thousand*Days)),
                          "k")) %>% 
  select(Year, `CC Point of Sale transactions per day`, `CC ATM withdrawals per day`)

dc_year <- debit_cards %>% 
  group_by(Year) %>% 
  summarise(`POS transactions` = sum(POSVOLUME),
            `POS Value` = sum(POSVALUE_TTM),
            `ATM transactions` = sum(ATMVOLUME),
            `ATM value` = sum(ATMVALUE_TTM))  %>% 
  left_join(ydl) %>% 
  mutate(`DC Point of Sale transactions per day` = paste0(round(`POS transactions`/(thousand*Days)),
                          "k"),
         `DC ATM withdrawals per day` = paste0(round(`ATM transactions`/(thousand*Days)),
                          "k")) %>% 
  select(Year, `DC Point of Sale transactions per day`, `DC ATM withdrawals per day`)

daily_volumes <- left_join(dc_year, cc_year)
daily_vol_tbl <- daily_volumes %>% 
  gt() %>% tab_spanner(
    label = "Credit cards",
    columns = vars(
      `CC Point of Sale transactions per day`, `CC ATM withdrawals per day`)
  ) %>% tab_spanner(
    label = "Debit cards",
    columns = vars(
      `DC Point of Sale transactions per day`, `DC ATM withdrawals per day`)
  ) %>% cols_align(
    align = "center",
    columns = vars(`DC Point of Sale transactions per day`, `DC ATM withdrawals per day`,
                   `CC Point of Sale transactions per day`, `CC ATM withdrawals per day`)
  ) %>% tab_style(
    style = list(
      cell_borders(
        sides = "right",
        color = "blue",
        weight = px(3)
      )
    ),
    locations = list(
      cells_body(
        columns = vars(Year, `DC ATM withdrawals per day`,
                       `CC ATM withdrawals per day`)
      )
    )
  )


daily_vol_tbl
```
```{r}
linx_debit_fee <- 0.75
credit_card_fee <- 0.02
debit_rev <- debit_cards %>% 
group_by(Year) %>% summarise(`POS transactions` = sum(POSVOLUME)) %>%
  mutate(`Debit card transaction revenue (TTM)` = round(linx_debit_fee*`POS transactions`/million))

credit_rev <- credit_cards %>%
  group_by(Year) %>% summarise(`POS Value (TTM)` = sum(POSVALUE_TTM)) %>% 
  mutate(`Credit Card transaction revenue (TTM)` = credit_card_fee*`POS Value (TTM)`)
```


```{r}




dc_narrow <- debit_cards %>% 
  select(DATE, ATMVALUE_TTM, POSVALUE_TTM) 

cc_narrow <- credit_cards %>% 
  select(DATE, POSVALUE_TTM) %>% 
  rename(CC_POS = POSVALUE_TTM)


payments_long <- left_join(dc_narrow, cc_narrow)  %>% 
  pivot_longer(c(ATMVALUE_TTM, POSVALUE_TTM, CC_POS), 
               names_to = "Measure", values_to = "Value")

```


```{r}
ggplot(data = payments_long,
       aes(x = DATE, y = Value, 
           group = Measure, color = Measure)) + 
  geom_line(size = 0.3) + geom_point() + 
  geom_smooth(se = FALSE) + 
  theme_ipsum() + 
  scale_y_continuous(labels = comma)
```


# Electronic Payment methods 

  * Payment cards
    * debit cards 
    * credit cards
  * ACH 
  * RTGS 
  
  
# Debit card

Most heavily-used form of electronic payment

Can be used for:
  1. Point-of-sale transactions
  2. Withdrawals from ATM machines
  
# Pre-paid cards

 Do not require an account. The card itself has the value attached to it. Can be used anonymously.
 
 Gift cards
 prepaid telephone cards
 
# ACH: Automated clearinghouse

Used for retail transactions: TT $500,000 or less)


# Telephone banking


# Mobile payments 



# The LINX network


 Government book grant cards


# Credit Cards



# Estimating Revenue at Point of Sales









