---
title: "Data visualisation with ggplot2"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    chakra: libs/remark.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5, warning = FALSE, message = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
duo_accent(
  primary_color = "ivory",
  secondary_color = "#310A31",
  header_font_google = google_font("Roboto", "400"),
  text_font_google   = google_font("Lato", "300"),
  code_font_family = "Fira Code",
  code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css",
  header_color = "#f54278",
  title_slide_text_color = "#354a66"
)
```

## ggplot2

  * popular data visualisation package
  
  * based on the **grammar** of **graphics** 
  

--

How do you access it? 

--

```{r}
library(ggplot2)
```

--

or

--

```{r eval = FALSE}
library(tidyverse)
```


---

## Underlying idea

Each graph/chart consists of three main components:

  * **data** 
  
--

  * an **aesthetic**: how are the variables in the data related to characteristics of the graph? E.g.:
  
--
    * x - variable determining values along a horizontal axis

    * y - variable determining values along a vertical axis
  
    * color/colour - variable determining the (external) colour of objects in the graph
--

  * a **geometry**: how is the mapping implemented geometrically i.e. what type of graph?
  
  
---
## Simplest ggplot2 graphs

The simplest ggplot2 graphs can be built with two objects:
  
  * a base ggplot object, using <tt>ggplot</tt>
  
  * a geometry object using one of the many <tt>geom_</tt> functions, e.g.:
    * <tt>geom_histogram</tt>
    * <tt>geom_boxplot</tt>
    * <tt>geom_bar</tt>
    * <tt>geom_point</tt>
    * <tt>geom_line</tt>
    
    
---
## Creating the aesthetic

Reminder: the aesthetic tells ggplot which variables to associate with specific characteristics of the graph.


--

The aesthetic is created using the <tt>aes</function>:




---
### Reading in the data


First, read in the data

```{r}
library(readr)
cars2018 <- read_csv("data/cars2018.csv")
```


---
### First graph: a histogram

Question: what does the distribution of MPG values look like? 

--

We can answer this using a histogram


--

Graph components: 

--

  * data: <tt>cars2018</tt>
--

  * aesthetic:
    * x = MPG
--
  * geometry: <tt>geom_histogram</tt>




---
### First graph: a histogram

```{r eval = FALSE }
ggplot(data = cars2018) 
```


---
### First graph: a histogram

```{r eval = FALSE }
ggplot(data = cars2018, aes(x = MPG)) 
```

---
### First graph: a histogram

```{r eval = FALSE }
ggplot(data = cars2018, aes(x = MPG)) + 
  geom_histogram()
```


---
### First graph: a histogram

```{r}
ggplot(data = cars2018, aes(x = MPG)) + 
  geom_histogram()
```


---
### First graph: a histogram


```{r}
ggplot(data = cars2018, aes(x = MPG)) + 
  geom_histogram(fill = "purple", colour = "black") 
```


---
### First graph: a histogram


```{r}
ggplot(data = cars2018, aes(x = MPG)) + 
  geom_histogram(fill = "purple", colour = "black") + theme_bw()
```

---
### First graph: a histogram


```{r}
ggplot(data = cars2018, aes(x = MPG)) + 
  geom_histogram(fill = "purple", colour = "black") + theme_bw() + 
  ggtitle("Distribution of Fuel Efficiency among 2018 Cars")
```


---
### New concept: faceting

Faceting allows us to see plots for specific levels of categorical variables using either the <tt>facet_wrap</tt> or <tt>facet_grid</tt> functions.

E.g. We can look at the distribution for each transmission type



---
### Faceted histogram


```{r}
ggplot(data = cars2018, aes(x = MPG)) + 
  geom_histogram(fill = "purple", colour = "black") + theme_bw() + 
  ggtitle("Distribution of Fuel Efficiency among 2018 Cars") + 
  facet_wrap(~Transmission, ncol = 1)
```



---
### A scatterplot:

What is the relationship between engine Displacement and fuel efficiency?

We can answer this using a scatterplot


--

Graph components: 

--

  * data: <tt>cars2018</tt>
--

  * aesthetic:
    * x = Displacement
    
    * y = MPG
--
  * geometry: <tt>geom_point</tt>



---
### Second graph: a scatterplot


```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG)) + 
  geom_point() 
```


---
### Second graph: a scatterplot


```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG, colour = Transmission)) + 
  geom_point() 
```

---
### Second graph: a scatterplot


```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG, colour = Transmission)) + 
  geom_point() + theme_bw() 
```



---
### Second graph: a scatterplot


```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG, colour = Transmission)) + 
  geom_point() + theme_bw() + 
  ggtitle("Fuel Efficiency vs Displacement")
```



---
### Dealing with overplotting 

We could try faceting

```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG)) + 
  geom_point() + theme_bw() + 
  ggtitle("Fuel Efficiency vs Displacement") +
  facet_wrap(~Transmission)
```

---
### Dealing with overplotting 

We could try increasing transparency

```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG)) + 
  geom_point(alpha = 0.3) + theme_bw() + 
  ggtitle("Fuel Efficiency vs Displacement") 
```


---
### Dealing with overplotting 

We could try jittering

```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG)) + 
  geom_jitter() + theme_bw() + 
  ggtitle("Fuel Efficiency vs Displacement") 
```


---
### Dealing with overplotting 

We could try jittering and increased transparency

```{r}
ggplot(data = cars2018, aes(x = Displacement, y = MPG)) + 
  geom_jitter(alpha = 0.5) + theme_bw() + 
  ggtitle("Fuel Efficiency vs Displacement") 
```



---
### Bar charts

How many incidents reported via 311 in each borough of NYC over the period Jan 1 2020 to Jan 14 2020?


First read in the data:

```{r}
nyc311 <- read_csv("data/nyc311.csv")
```


---
### Using a bar chart  

We can show the number of calls in each borough using a bar chart

  * data: <tt>nyc311</tt>

  * aesthetic: 
    * x: Borough
    
    
  * geometry: <tt>geom_bar</tt>
    * note that the default behavior of <tt>geom_bar</tt> is to use the count, so y aesthetic is not necessary. 
  
  
---
### Bar chart of NYC 311 calls by borough


```{r}
ggplot(data = nyc311, aes(x = Borough)) + geom_bar()
```




---
### A bar chart with sums instead of counts

How many runs did Brian Lara score in test matches against each test-playing country?

First, get the data:

```{r}
lara <- read_csv("data/lara.csv") 
```

---
### A bar chart with sums instead of counts

How many runs did Brian Lara score in test matches against each test-playing country?


First get the data:

```{r}
library(dplyr)
library(tidyr)
lara <- read_csv("data/lara.csv") %>% 
  separate(Match, into = c("Type", "Number"))

tests <- filter(lara, Type == "Test")
```


---
### Now to make the chart

  * data: <tt>tests</tt>
  
  * aesthetic:
    * x: Opp
    * y: Runs
    
  * geometry: <tt>geom_bar</tt>
    * now we have to change the statistic
    
    
---
### Making the bar chart

```{r}
ggplot(data = tests, aes(x = Opp, y = Runs)) + 
  geom_bar(stat = "identity")
```

