---
title: "Subsetting data using R"
subtitle: "📈📊📉"
author: "Daryn Ramsden"
institute: ""
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```
### A note 

These slides assume your working directory contains the *data* folder that contains the following files:
  - *Loans.csv*
  - *ChicagoEmployees.csv*

One way to ensure that you are in that directory is to open the *week2.Rproj* file. (After downloading the project folder *week2* from elearning.)

You can do this in RStudio by File --> Open Project and then navigating to it at the location you downloaded it to.


---
### Reading in data

We will use the **read_csv** function of the <tt>readr<package></tt>. 

**read_csv** is an alternative to the built-in function **read.csv**. 

It has some smarter default settings than read.csv

Also, it gives info about the names and data types of columns in the data set (which can sometimes be annoying) 

---
#### Using **read_csv** 


```{r}
#install.packages("readr") # you may need to run this first if you ha
library(readr)
loans <- read_csv("data/Loans.csv")
```


---
### Getting summary information

How big is the loans data set? 
--
```{r}
dim(loans)
```
--
What are the names of the columns?
```{r}
names(loans)
```
--
You can use **summary** or **View** as well to get more info. 



---
### Selecting only some columns

In base R, **[]** is used to subset data. For a 2d data structure, we have row conditions and column conditions separated by a comma inside the **[]**.
--


For example, we can select the first 5 rows and the first 4 columns of the *loans* data frame:

--
```{r}
loans[c(1:5),c(1:4)]
```



---
### Selecting columns by name

It's often more practical to select columns by name however. The following code:
  - keeps only the *homeownership*, *loan_amount*, *loan_purpose*, and *interest_rate* columns
  - displays the dimensions of the smaller data set
--
```{r}
loans2 <- loans[, c("homeownership", "loan_amount", 
                    "loan_purpose", "interest_rate")]
dim(loans2)
```
--
Note that there was no row constraint in the first line as we were intending to keep all rows of the data.

---
### Rows can also be selected using logical operators

How do we select all the rows where the interest rate is lower than 7%?
--


The code below:
  - keeps only the rows where the interest rate is lower than 7% 
  - displays the first few rows
--
```{r}
low_interest <- loans2[loans2$interest_rate < 7, ]
head(low_interest)
```
--
Note that there is no column constraint as we are keeping all the columns.


---
### What are the other logical operators that can be used in filtering rows?

The previous slide made use of **<**, multiple other operators are available for filtering: 
--
```{r echo = FALSE}
Operator <- c( "==", "!=", "<", "<=",  ">", ">=", "|", "&", "!")
Meaning <- c("equal", "not equal", 
             "less than", "less than or equal to", 
              "greater than", "greater than or equal to",
             "Or: at least one of the expressions is true", 
             "And: Both expressions are true", 
             "Not: the expression is not true" )
optable <- data.frame(Operator, Meaning)
kableExtra::kable(optable)
```




---
### Some examples 

How would we get loans that meet both of the following:
  - have a loan amount > $20,000
  - interest rate > 12% 
--

```{r}
df <- loans2[(loans2$loan_amount > 20000) & 
               (loans2$interest_rate > 12), ]
head(df)
```


---
### Another example

How would we get loans that meet either of the following:
  - the purpose is *credit_card* 
  - the home ownership status is *RENT* 
  
--
```{r}
df2 <- loans2[loans2$loan_purpose == "credit_card" | 
                loans2$homeownership == "RENT", ] 
head(df2)
```

  

---
### Using **$** to reference columns 

Another means of subsetting data that we would have seen before is the **$** 
--
```{r fig.height=5}
hist(loans$interest_rate)
```



--


We have been using it to extract single columns from the data


---
### **$** can be used to create new columns

We can also create new columns in the data set. 


For example, here's a new column that says whether or not the interest rate on a loan is greater than 10%:
--
```{r}
loans2$ir_gt10 <- ifelse(loans2$interest_rate > 10, TRUE, FALSE)
head(loans2)
```


---
### Explaining **ifelse**

The **ifelse** function is very handy and has three arguments:

- an expression to evaluate
- a result if the expression is TRUE
- a result if the expression is FALSE

---
### dplyr: An alternative means for subsetting data 

Subsetting in base R can involve a lot of typing.

--

The *dplyr* package contains useful data manipulation functions to addresss this. 

--

Installing **dplyr** 

--
```{r eval = FALSE}
install.packages("dplyr")
```
--

Now getting access to the functions:
```{r message = FALSE, warning = FALSE}
library(dplyr)
```



---
### 5 main dplyr functions

There are 2 main functions we will use for subsetting data:
- **filter**: to choose rows
- **select**: to choose columns

Other functions, we will introduce: 
- **arrange**: to sort
- **mutate**: to create new columns
- **summarise** and **group_by**: to aggregate data 


---
### General rules about usage 

Key things to remember: 
- **the first argument/input to any function in dplyr, is the data frame** 


- the other inputs depend on the task you're carrying out  

---
### Example: Using dplyr::filter to select rows


How would we get loans that meet either of the following:
  - the purpose is *credit_card* 
  - the home ownership status is *RENT* 
--
```{r}
library(dplyr)
df4 <- filter(loans2, loan_purpose == "credit_card" | 
                homeownership == "RENT")
head(df4)
```

---
### Example 2: Using dplyr::filter to select rows

How would we get loans that meet both of the following:
  - have a loan amount > $20,000
  - interest rate > 12% 

--
```{r}
library(dplyr)
df5 <- filter(loans2, loan_amount > 20000, interest_rate > 12)
head(df5)
```


---
### Example 3: Using dplyr::select to select certain columns

Going back to the original *loans* data set, we will select only the following columns:
- *homeownership*
- *loan_amount*
- *loan_purpose*
- *interest_rate*
--
```{r}
df6 <- select(loans, homeownership, loan_amount, 
                    loan_purpose, interest_rate)
head(df6)
```


---
### Example 4: Using dplyr::mutate to add columns

Now to add a column indicating if the interest rate is greater than 10% using **mutate**
--
```{r}
df7 <- mutate(df6, ir_gt10 = ifelse(interest_rate > 10, 
                                    TRUE, FALSE))
head(df7)
```



---
### Example 5: Using dplyr::arrange to sort the data

If we want to sort by *loan_amount*
--
```{r}
df8 <- arrange(df7, loan_amount)
head(df8)
```

--


The default sort is in ascending order 


---
### Example 6: sorting data in descending order

We can set multiple columns to sort by (and sort in descending order as well):
--
```{r}
df9 <- arrange(df7, desc(loan_amount), desc(interest_rate))
```


---
### Aggregating data 

Data aggregation in *dplyr* typically involves using the **group_by** and **summarise** functions.
--
- First use group_by to specify how you want to group the data 


- apply summarise to aggregate the data

---
### Example 7: Using summarise and group_by

First, we group the data
--
```{r}
grouped_data <- group_by(df7, loan_purpose)
head(grouped_data)
```
--
The grouped data frame looks the same but R has added extra information that will be used by **summarise** (or other functions)


---
### Example 7 continued: summarising grouped data

Now we can use **summarise** with the grouped data:
--
```{r}
agg_data <- summarise(grouped_data, total_amt = sum(loan_amount), 
                      num_loans = n(),
                      high_int_pct = mean(ir_gt10)*100)

agg_data
```

---
### Example 7 continued: sorting the aggregated data 


And what if we wanted to order the loan types by the total loan amount for each? 

--
```{r}
sorted_data <- arrange(agg_data, desc(total_amt))
sorted_data

```









